---
title: "Cirrhosis parametric survival model"
date: "last-modified"
format: html
resource_files:
- cirrhosis.html
---

# About data

The data provided is sourced from a Mayo Clinic study on the randomised controlled trial for the drug D-penicillamine for primary biliary cirrhosis (PBC) of the liver carried out from 1974 to 1984. 312 patients took part in the study, 112 patients did not join the trial but agreed to record basic metrics and survival tracking but 6 of them soon went untraceable, leaving 102 of them. The dataset encompasses information on 418 patients followed over a period denoted by N_Days. This timeframe tracks the duration from patient registration until either death, liver transplantation, or the study's conclusion (July 1986). It incorporates the data on patients' sociodemographic characteristics, medical history, clinical signs, disease severity and outcome. There was 17 clinical features available for predicting survival state of patients with liver cirrhosis. The survival states include 0 = D (death), 1 = C (censored), 2 = CL (censored due to liver transplantation).

# Data Analysis

## Exploratory data analysis

### Load environment

```{r}
library(flexsurv)
library(survival)
library(tidyverse)
library(gtsummary)
library(survminer)
data0 <- read.csv('cirrhosis.csv')

```

### Explore Data

```{r}
summary(data0)
```

### Data wrangling

set outcome "N_days" as numeric data

```{r}
data1 <- data0 %>%
  mutate(N_Days = as.numeric(N_Days))
```

Clean and prepare data

```{r}
data1 <- data1 %>%
  mutate(
    status_binary = recode(Status, "D" = 1, "C" = 0, "CL" = 0),
    status_binary = factor(status_binary, levels = c(0, 1), labels = c("censored", "death")),
    Sex = recode(Sex, "F" = "Female", "M" = "Male"),
    Sex = as.factor(Sex),  # Ensure sex is a factor
    Bilirubin = as.numeric(Bilirubin),  # Ensure Bilirubin is numeric
    time = as.numeric(N_Days),  # Use N_days (now numeric) for survival analysis
    age_years = as.numeric(duration(Age, units = "days") / dyears(1)),  # Convert age in days to age in years using lubridate
    Stage = as.factor(Stage)  # Ensure Stage is a factor
  ) %>%
  mutate(across(where(is.character), ~ recode(., "N" = "No", "Y" = "Yes"))) %>%
  mutate_if(is.integer, as.numeric) %>%  # Convert all integer columns to numeric
  mutate_if(is.character, as.factor) %>%  # Convert all character columns to factors
  filter(!is.na(time), !is.na(Bilirubin), !is.na(Sex), !is.na(age_years), !is.na(Stage))
```

```{r}
str(data1)
```

```{r}
summary(data1)
```

### Describe data

overall summary table

```{r}
summary_overall <- data1 %>%
  select(-ID, -Status, -Age) %>%
  tbl_summary(
    by = status_binary,
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
     ) %>%
      add_overall() %>% 
      add_p()
summary_overall
```

## **Kaplan-Meir Survival estimates**

### **Kaplan-Meier estimates for overall**

```{r}
KM1 <- survfit(Surv(time = data1$time, event = data1$status_binary == "death") ~ 1, 
               type = "kaplan-meier", data = data1)
summary(KM1)
```

Plot the survival probability

```{r}
ggsurvplot(KM1, data = data1, risk.table = TRUE, linetype = c(1,2), pval = TRUE)
```

### **Kaplan-Meier estimates for groups**

```{r}
KM1.g <- survfit(Surv(time = data1$time, event = data1$status_binary == "death") ~ Sex, 
               type = "kaplan-meier", data = data1)
summary(KM1.g)
```

```{r}
ggsurvplot(KM1.g, data = data1, risk.table = TRUE, 
           linetype = c(1,2), pval = TRUE)
```

```{r}
KM1.stage <- survfit(Surv(time = data1$time, event = data1$status_binary == "death") ~ Stage, 
               type = "kaplan-meier", data = data1)
summary(KM1.stage)
```

```{r}
ggsurvplot(KM1.stage, data = data1, risk.table = TRUE, 
           linetype = c(1,2,3,4), pval = TRUE)
```

comparing Kaplan Meier across groups

```{r}
survdiff(Surv(time = data1$time, event = data1$status_binary == "death") ~ Sex, 
         data = data1, 
         rho = 0)

```

No significant difference in survival estimates between the male and female by log rank test (p=0.08)

```{r}
survdiff(Surv(time = data1$time, event = data1$status_binary == "death") ~ Stage, 
         data = data1, 
         rho = 0)

```

There was a statistically significant difference in survival distributions between the groups of stages .

### Kaplan Meier Cumulative Hazard Function Estimate

```{r}
ggsurvplot(
  KM1,
  fun = "cumhaz",
  pval.method = TRUE,
  conf.int = TRUE,
  ggtheme = theme_light(),
  title = "Kaplan-Meier Cumulative Hazard Function Estimate"
)
```

```{r}
epiR::epi.insthaz(KM1) %>%
  ggplot(aes(x = time, y = hest)) + 
  geom_smooth(color = "red", method = "loess", formula = "y ~ x") +
  theme_light() +
  labs(title = "Kaplan-Meier Hazard Function Estimate", 
       x = "Time", y = "Instantaneous Hazard")
```

## Fitting the Parametric Survival Model

Parametric survival model able to describe the effect of multiple covariates on survival of PCB patient, assuming the survival time follows parametric distribution. Clinically significant covariates such as age, stage of cirrhosis and serum bilirubin are included to predict survival in the following parametric survival models.

### Exponential model

In the exponential survival model, the hazard is assumed to be constant over time.

```{r}
surv.mod <- Surv(time = data1$time, event = data1$status_binary == "death")
exp.mod <- survreg(surv.mod ~ age_years + Stage + Bilirubin, data = data1, dist = "exponential")
```

```{r}
summary(exp.mod)
```

```{r}
tbl_regression(exp.mod)
```

Interpretation

1.Â Â Â Â Â  Age

-   The estimated log time ratio to die for every increment of 1 year of age is -0.03. The Acceleration Factor (AF) or Time Ratio (TR) is ð‘’^âˆ’0.03^=0.972. This indicate with the addition of each year, the time to death is accelerated by a factor of 0.972.

-   The estimated hazard ratio (HR) with addition of one year is e^0.03^=1.029, which means the risk of death is significantly higher by 2.9% with the increase of one year of age when adjusted for the stage of cirrhosis and serum bilirubin.

2.Â Â Â Â Â  stage

-   The estimated log time ratio to die in PCB patient in stage 4 in comparison to stage 1 is -1.878. The Acceleration Factor (AF) or Time Ratio (TR) is ð‘’^âˆ’1.878^=0.153, meaning being in Stage 4 accelerates the time to death by a factor of 0.153 compared to Stage 1.

-   The estimated hazard ratio (HR) comparing Stage 4 to Stage 1 is e^1.878^=6.538, which means patient in Stage 4 have 6.5 times higher risk of death compared to those in Stage 1 when adjusted for the age and serum bilirubin.

3.Â Â Â Â Â  Bilirubin

-   The estimated log time ratio to die for every increment of 1mg/dl of serum bilirubin is -0.127 The Acceleration Factor (AF) or Time Ratio (TR) is ð‘’^âˆ’0.127^=0.881. This indicate with 1mg/dl increase in serum bilirubin, the time to death is accelerated by a factor of 0.881.

-   The estimated hazard ratio (HR) comparing the increase of 1mg/dl of serum bilirubin is e^0.127^=1.135, which means the risk of death is higher by 13.5% with each 1mg/dl increase in serum bilirubin when adjusted for the stage of cirrhosis and age.

### Weibull model

#### Accelerated Failure Time (ATF) model

```{r}
wei.mod.aft <- flexsurvreg(surv.mod ~ age_years + Stage + Bilirubin , data = data1, dist = 'weibull')
wei.mod.aft
tbl_regression(wei.mod.aft)
```

The weibull scale parameter is 0.728 and the shape parameter is 1/0.728 = 1.37.

Interpretation

1.Â Â Â Â Â  Age

-   The estimated log time to death wtih the increase in one year of age was -0.022

-   The Acceleration Factor (AF) or Time Ratio (TR) is ð‘’^âˆ’0.022^=0.979. This indicate with the addition of each year, the time to death is accelerated by a factor of 0.979. in other words, the older individual has shorter time (by a factor of 0.979) to death when adjusted for the serum bilirubin and stage of cirrhosis.

-   The reduction in survival in PCB patients could range between 1% to 3.3% (adj.TR= 0.979, 95% CI= 0.967, 0.990)

2.Â Â Â Â Â  stage

-   The estimated log time to death for PCB patient in stage 4 in comparison to Stage 1 was -1.457.
-   The Acceleration Factor (AF) or Time Ratio (TR) is ð‘’^âˆ’1.457^=0.233, meaning being in Stage 4 accelerates the time to death by a factor of 0.233 compared to Stage 1 when adjusted for the serum bilirubin and age.
-   The survival in PCB patients shortened between 34.2% to 91.7% (adj.TR= 0.233, 95% CI= 0.083, 0.658)

3.Â Â Â Â Â  Bilirubin

-   The estimated log time to death wtih the increase 1mg/dl of serum bilirubin was -0.105
-   The Acceleration Factor (AF) or Time Ratio (TR) is ð‘’^âˆ’0.105^=0.900. This indicate with 1mg/dl increase in serum bilirubin, the time to death is accelerated by a factor of 0.9. Higher serum bilirubin level are associated with shorter survival time by a factor of 0.9 when adjusted for the stage of cirrhosis and age.
-   The reduction in survival in PCB patients could range between 8.4% to 11.6% (adj.TR= 0.900, 95% CI= 0.884, 0.916)

#### Proportional Hazard (PH) model

```{r}
wei.mod.ph <- flexsurvreg(surv.mod ~ age_years + Stage + Bilirubin, data = data1, dist = "weibullPH")
wei.mod.ph
tbl_regression(wei.mod.ph)
```

Interpretation

1.Â Â Â Â Â  Age

-   The estimated hazard ratio (HR) with addition of one year is e^0.030^=1.03, which means the risk of death is higher by 3% with the increase of one year of age when adjusted for the serum bilirubin and stage of cirrhosis.
-   The risk of death in PCB patients could increase by the range between 1% to 5% (adj.TR= 1.03, 95% CI= 1.01, 1.05 )

2.Â Â Â Â Â  stage

-   The estimated hazard ratio (HR) comparing Stage 4 to Stage 1 is e^2.00^=7.40, which means patient in Stage 4 have 7.4 times higher risk of death compared to those in Stage 1 when adjusted for the serum bilirubin and age.

3.Â Â Â Â Â  Bilirubin

-   The estimated hazard ratio (HR) comparing the increase of 1mg/dl of serum bilirubin is e^0.145^=1.16 which means the risk of death is higher by 16% with each 1mg/dl increase in serum bilirubin when adjusted for the stage of cirrhosis and age.
-   the risk of death increases ranging 13% to 18% (adj.TR= 1.16, 95% CI= 1.13, 1.18 )

## **Model adequacy for Weibull distribution**

**Weibull regression diagnostic plot by Stage**

```{r}
library(SurvRegCensCov)
diagnostic_plot <- WeibullDiag(Surv(time = data1$time, event = data1$status_binary == "death") ~ Stage, data = data1)

```

**Weibull regression diagnostic plot by sex**

```{r}
diagnostic_plot <- WeibullDiag(Surv(time = data1$time, event = data1$status_binary == "death") ~ Sex, data = data1)
```

Weibull model is fit, the groups' lines are parallel and linear (not flat).

## Checking the PH assumption

Plot of log-odds of survival against the log of survival time by sex

```{r}
kmfit <- survfit(surv.mod ~ data1$Stage)
plot(log(kmfit$time), log(kmfit$surv/(1-kmfit$surv)))

```

Plot of log-odds of survival against the log of survival time by sex

```{r}
kmfit1 <- survfit(surv.mod ~ data1$Sex)
plot(log(kmfit1$time), log(kmfit1$surv/(1-kmfit1$surv)))

```

The plots do not look like a straight lines. It is also not parallel. So it does not support PH assumption. OurÂ **PH asssumption is violated**.

## Comparing models

```{r}
exp.mod_AIC <- AIC(exp.mod)
wei.mod.aft_AIC <- wei.mod.aft$AIC
wei.mod.ph_AIC <- wei.mod.ph$AIC
```

```{r}
aic_values <- data.frame(
  Model = c("Weibull AFT", "Weibull PH", "Exponential"),
  AIC = c(wei.mod.aft_AIC, wei.mod.ph_AIC, exp.mod_AIC)
)

print(aic_values)
```

Weibull model is a better fit model compared to exponential model as it has lower AIC.

## Plot Survival Probability

```{r}
weib.cirr <- survreg(Surv(time, status_binary == "death") ~ Bilirubin, data = data1, dist = 'weibull')
summary(weib.cirr)
```

Weibull survival curve for a cirrhosis subject with bilirubin of 8mg/dl

```{r}
pct <- 1:98/100 # The 100th percentile of predicted survival is at +infinity
ptime <- predict(weib.cirr, newdata=data.frame(Bilirubin = 8) ,
type = 'quantile', p = pct, se = TRUE)
matplot(cbind(ptime$fit, ptime$fit + 2*ptime$se.fit,
ptime$fit - 2*ptime$se.fit)/30.5, 1-pct,
xlab = "Days", ylab = "Survival", type = 'l', lty = c(1,2,2), col = 1)
```

# **Prepared By**

1.Â Â Â Â Â  Tiong Wen Ning 22307078

2.Â Â Â Â Â  Hazlienor Binti Mohd Hatta 22307055

3.Â Â Â Â Â  Amalina Binti Asmail \@ Ismail Â  22307044

4.Â Â Â Â Â  Mohamad Afiq Amsyar Bin Hamedin 23100003

5.Â Â Â Â Â  Muhammad Saufi bin AbdullahÂ  22307048
