---
title: "Cirrhosis parametric survival model"
date: "last-modified"
format: html
resource_files:
- cirrhosis.html
---

# About data

The data provided is sourced from a Mayo Clinic study on primary biliary cirrhosis (PBC) of the liver carried out from 1974 to 1984. The dataset encompasses information on 418 patients followed over a period denoted by N_Days. This timeframe tracks the duration from patient registration until either death, liver transplantation, or the study's conclusion (July 1986). It incorporates the data on patients\' sociodemographic characteristics, medical history, clinical signs, disease severity and outcome. Utilize 17 clinical features for predicting survival state of patients with liver cirrhosis. The survival states include 0 = D (death), 1 = C (censored), 2 = CL (censored due to liver transplantation).

# Data analysis

## Exploratory data analysis

### Load environment

```{r}
library(flexsurv)
library(survival)
library(tidyverse)
library(gtsummary)
data1 <- read.csv('cirrhosis.csv')

```

### Data wrangling

rename outcome into time as set as numeric data

```{r}
data1$time <- data1$N_Days
data1$time <- as.numeric(data1$time)
```

set all character as factor

```{r}
data1[] <- lapply(data1, function(x) {
if (is.character(x)) {
factor(x)
} else {
x
}
})
```

check data structure

```{r}
str(data1)
```

recode event of interest "D" as "1", while "C" and "CL" as "0"

```{r}
library(dplyr)
data1 <- data1 %>%
mutate(status_binary = recode(Status, "D" = 1, "C" = 0, "CL" = 0))
```

### Describe data

```{r}
data1 %>%
select(-ID, -N_Days, -Status) %>%
tbl_summary(by = status_binary,
statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
add_overall() 
```

## Fitting the Parametric Survival Model

### Exponential model

```{r}
surv.mod <- Surv(time = data1$time, event = data1$status_binary == 1)
exp.mod <- survreg(surv.mod ~ Sex + Bilirubin + Stage + Prothrombin , data = data1, dist='exponential')
summary(exp.mod)
tbl_regression(exp.mod)
```

Interpretation

The exponential model assumes a constant hazard rate over time with the scale parameter fixed at 1.

1.Â Â Â Â Â  Sex

oÂ Â  Males have a survival time ð‘’^âˆ’0.5167^=0.596, that is 59.6% of the survival time for females, meaning males have shorter survival times compared to females.

oÂ Â  The hazard for males is e^0.5167^=1.676, male is 1.676 times higher than for females, indicating a higher risk of the event (death) for males.

2.Â Â Â Â Â  Bilirubin

oÂ Â  For each unit increase in bilirubin, the expected survival time decreases to 89.1% (e^âˆ’0.1151^=0.891) of the previous level, indicating shorter survival times with higher bilirubin levels.

oÂ Â  The hazard increases by ð‘’ ^0.1151^=1.122, 12.2% for each unit increase in bilirubin, indicating a higher risk of death with higher bilirubin levels.

3.Â Â Â Â Â  stage

oÂ Â  Patients in a higher stage have a survival time e^âˆ’0.6356^ that is 53% of that of patients in the lower stage, indicating much shorter survival times with advanced disease stages.

oÂ Â  The hazard is e ^0.6356^=1.888, 1.888 times higher for each unit increase in stage, indicating a significantly higher risk of death with more advanced disease stages.

4.Â Â Â Â Â  prothrombin

oÂ Â  For each unit increase in prothrombin, the survival time decreases to e^âˆ’0.2578^=0.772, 77.2% of the previous level, indicating shorter survival times with higher prothrombin levels.

oÂ Â  The hazard increases by e ^0.2578^=1.294, 29.4% for each unit increase in prothrombin, indicating a higher risk of death with higher prothrombin levels.

### Weibull model

#### Accelerated Failure Time (ATF) model

```{r}
wei.mod <- survreg(surv.mod ~ Sex + Bilirubin + Stage + Prothrombin, data = data1, dist = 'weibull')
summary(wei.mod)
tbl_regression(wei.mod)
```

Interpretation

1.Â Â Â Â  Sex

oÂ Â  Males have a survival time ð‘’^âˆ’0.39336^=0.675, that is 67.5% of the survival time for females, meaning males have shorter survival times compared to females.

2.Â Â Â Â  Bilirubin

oÂ Â  For each unit increase in bilirubin, the expected survival time decreases to 90.6% (e^âˆ’0.09891^=0.906) of the previous level, indicating shorter survival times with higher bilirubin levels.

3.Â Â Â Â  stage

oÂ Â  Patients in a higher stage have a survival time e^âˆ’0.51456^ that is 59.8% of that of patients in the lower stage, indicating much shorter survival times with advanced disease stages.

4.Â Â Â Â  prothrombin

oÂ Â  For each unit increase in prothrombin, the survival time decreases to e^âˆ’0.18369^=0.832, 83.2% of the previous level, indicating shorter survival times with higher prothrombin levels.

#### Proportional Hazard  (PH) model

```{r}
wei.mod.ph <- flexsurvreg(Surv(time, status_binary) ~ Sex + Bilirubin + Stage + Prothrombin, data = data1, dist = 'weibullPH')
wei.mod.ph
tbl_regression(wei.mod.ph)
```

##### Interpretation

1.Â Â Â Â  Sex

oÂ Â The hazard for males is e^0.544^=1.72, male is 1.72 times higher than for females, indicating a higher risk of the event (death) for males.

2.Â Â Â Â  Bilirubin

oÂ Â  The hazard increases by ð‘’ ^0.137^=1.15, 15% for each unit increase in bilirubin, indicating a higher risk of death with higher bilirubin levels.

3.Â Â Â Â  stage

oÂ Â  The hazard is e ^0.711^=1.888, 2.04 times higher for each unit increase in stage, indicating a significantly higher risk of death with more advanced disease stages.

4.Â Â Â Â  prothrombin

oÂ Â  The hazard increases by e ^0.254^=1.29, 29% for each unit increase in prothrombin, indicating a higher risk of death with higher prothrombin levels.

## **Model adequacy for Weibull distribution**

**Weibull regression diagnostic plot by sex**

```{r}
library(SurvRegCensCov)
diagnostic_plot <- WeibullDiag(Surv(time = data1$time, event = data1$status_binary == 1) ~ Sex, data = data1)

```

**Weibull regression diagnostic plot by stages of cirrhosis**

```{r}
diagnostic_plot <- WeibullDiag(Surv(time = data1$time, event = data1$status_binary == 1) ~ Stage, data = data1)
```

Weibull model is fit, the groups' lines are parallel and linear (not flat).

## Checking the PH assumption

Plot of log-odds of survival against the log of survival time by sex

```{r}
kmfit <- survfit(surv.mod ~ data1$Sex)
plot(log(kmfit$time), log(kmfit$surv/(1-kmfit$surv)))

```

Plot of log-odds of survival against the log of survival time by stage of cirrhosis

```{r}
kmfit1 <- survfit(surv.mod ~ data1$Stage)
plot(log(kmfit1$time), log(kmfit1$surv/(1-kmfit1$surv)))

```

The plots do not look like a straight lines. It is also not parallel. So it does not support PH assumption. OurÂ **PH asssumption is violated**.

## Plot Survival Probability

```{r}
weib.cirr <- survreg(Surv(time, status_binary) ~ Bilirubin, data = data1, dist = 'weibull')
summary(weib.cirr)
```

Weibull survival curve for a cirrhosis subject with bilirubin of 8mg/dl

```{r}
pct <- 1:98/100 # The 100th percentile of predicted survival is at +infinity
ptime <- predict(weib.cirr, newdata=data.frame(Bilirubin = 8) ,
type = 'quantile', p = pct, se = TRUE)
matplot(cbind(ptime$fit, ptime$fit + 2*ptime$se.fit,
ptime$fit - 2*ptime$se.fit)/30.5, 1-pct,
xlab = "Days", ylab = "Survival", type = 'l', lty = c(1,2,2), col = 1)
```

# **Prepared By**

1.Â Â Â Â Â  Tiong Wen Ning Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â  22307078

2.Â Â Â Â Â  Hazlienor Binti Mohd HattaÂ Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â  22307055

3.Â Â Â Â Â  Amalina Binti Asmail \@ Ismail Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â Â  22307044

4.Â Â Â Â Â  Mohamad Afiq Amsyar Bin Hamedin Â Â Â  23100003

5.Â Â Â Â Â  Muhammad Saufi bin Abdullah Â Â Â Â Â Â Â Â Â Â Â Â Â Â  22307048
